Assessing the Global COVID19 Impact on Air Transport with Open Data

### 2021-07-12 paper.Rmd
Covid data included

---
title: "Assessing the Global COVID19 Impact on Air Transport with Open Data"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  bookdown::pdf_book:
     base_format: rticles::ieee_article
bibliography: "DASC2021.bib"
affiliation:
  ## use one only of the following
  # author-columnar: true         ## one column per author
  institution-columnar: true  ## one column per institution  
  # wide: true          ## one column wide author/affiliation fields
  institution:
    - name: EUROCONTROL
      department: Performance Review Unit
      location: Brussels (Belgium)
      #mark: 1
      author:
        - name: Rainer Koelle
          email: rainer.koelle@eurocontrol.int
    - name: DECEA
      department: Subdepartment of Operations 
      location: Rio de Janeiro (Brazil)
      #mark: 2
      author:
        - name: Fabio Lourenco Carneiro Barbosa
          email: barbosaflcb@fab.mil.br
abstract: |
  This paper approaches the impact of the pandemic as a massive service disruption of the pre-pandemic global     connectivity and regional air transport networks. In particular, the project aims to provide data analytical evidence for policy success and transformation of the air transportation system. As an aspirational goal, the industry aims to recover in a “greener” manner. The project builds on openly available data sets. The paper will be produced in a reproducible manner making the data, code, and its processing available to interested reseachers and practitioners. The open assessment will provide policy makers with a tool to assess the reaction to local or regional measures.
  <!-- motivation -->
  <!-- short background - why important -->
  <!-- approach chosen / research method -->
  <!-- experimental work -->
  <!-- results obtained -->
  <!-- conclusions / key take-away --> 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(stats)
library(zoo)
```

```{r globalstuff, message=FALSE}
## set bookdown specs/defaults for high quality output
#---------- check for the settings

@@

###Preparatory codes

#If someone needs filters below, here we can control any selective sample
#Filtering Brazilian and European data to reduce the sample (and csv) size
#Currently, those filters are not in use in the code, they are here just in case.
bra_10_apts <- c("SBBR", "SBGR", "SBSP", "SBKP", "SBRJ", "SBGL", "SBCF", "SBSV", "SBPA", "SBCT")
eur_apts <- c("EHAM","LFPG","EGLL","EDDF","EDDM","LEMD","LIRF","LEBL","EGKK","LSZH")
study_airports <- c(bra_10_apts, eur_apts)

###Preparing airport file
#NOTE: the airports.csv file below is the downloaded file from www.ourairports.org.
airports <- read_csv("data-raw/airports.csv") %>% mutate(continent = case_when(is.na(continent) ~ "NA", TRUE ~ continent))

#There are missing airports and too much variables
apt_countries <- airports %>% transmute(ICAO = ident, CTRY = iso_country) %>% add_row(ICAO  = c("SPJC", "YBMC", "LSZM", "YSCH", "EPLB", "K3M3", "VV01", "SC28", "CWF2", "EHDB", "74XA", "HE13"), CTRY  = c("PE", "AU", "FR", "AU", "PL", "US", "VN", "US", "CA", "NL", "US", "EG"))

#If you need to write
#write_csv(apt_countries, "./data/apt_countries.csv")
#apt_countries is ready.

# Associate the regions
a <- airports %>% filter(continent == "EU") %>% select(iso_country) %>% unique()
eur_countries <- a$iso_country
#eur_countries is ready.

```

```{r}
#CURRENT COMMENTS AND TO-DO'S
# I HAVE DOWNLOADED 3 FILES FOR NOW (APR/19, APR/2020, APR/2021), JUST TO START "TIDYING" AND EXPLORING.
#IT'S IN THE DATA-RAW FOLDER (NOT SHARED WITH GITHUB), AS ALWAYS.
```

@@

However, the recent COVID-19 crisis posed a completely different, unexpected, and inverted challenge. Demand for air transport dropped as low as 90% of the previous "normal" in many places. Where the lack of capacity was previously the issue, now the lack of demand threatened the ATM system stability. In the financial perspective, airlines and airports had to deal with an unprecedented decrease in incomes. As a result, air navigation providers collected less fees for their services, due to significantly fewer flights. In the operational perspective, pilots and air traffic controllers practiced less. The problems and obstacles developed into many other dimensions.

Hence, the current scenario is a proper moment to further investigate the concept of resilience.

@@

# # Purpose Statement
# 
# ???The purpose of this research is to 
# investigate additional dimensions in which 
# resilience could be measured, in addition
# to the current framework of delay analysis.
# 
# Research Question
# 
# ???How can we enlarge the concept of 
# resilience, so that it is applicable to
# scenarios of low traffic? 
#
# ???Research Question: 
# ???RQ1.What was the impact of the pandemic on ATM resilience?
# ???  RQ1.1 How resilience can be modeled in a low-demand scenario?
# ???  RQ1.2 How resilient were different ATM systems worldwide?

```

@@

@@

-   data-analytical approach - using open data / freely available (tbd: validated against organisational data)
-   ???RQ1.1 = through a qualitative analysis of previous proposed models
-   ???RQ1.2 = through a quantitative analysis of open data

@@

@@

@@

@@

@@

This study builds on publicly available data. Opensky Network collects crowdsourced air traffic data from more than 2500 feeders (sensor stations). To support the process of illustrating and studying the impact of the COVID pandemic on air traffic demand, a flight-by-flight dataset is provided on a monthly basis [@xavier_olive_2021_4893103]. The data spans the period since 1. January 2019. Fig. \@ref(fig:osndaily) shows the number of daily flights tracked by Opensky Network globally.

@@

```{r osndaily, fig.cap="(ref:osndaily)", message=FALSE}
daily_tfc <- read_csv("./data/daily_osn.csv")

daily_tfc %>%
  ggplot(mapping = aes(x = DATE, y = FLTS)) +
  geom_line() +
  labs(x = NULL, y = "flights")
```

@@

```{r}
#Reading raw data

source("./R/list_apt_files.R")

#Here I will assign only one month - "202105". If you want to include a full year, just assign year to "2021" or "2020". It works.
year <- "2021"
file_names <- list_apt_files(.year = year)
open_sky <- map_dfr(file_names, read_csv)

#Selecting relevant variables

fb <- open_sky %>% transmute(ADEP = origin, ADES = destination, TYPE = as.factor(typecode), DATE = date(day), CALL = callsign
                             #, ACFT_ID = aircraft_uid
                             )

#Easily "dropping NA's# - this can be further sofisticated

fb1 <- fb %>% drop_na()
fb1
```

```{r}
#Joining to ADEP
fb2 <- left_join(fb1, apt_countries, by = c("ADEP" = "ICAO")) %>% mutate(ADEP_CTRY = CTRY, .keep = "unused", .after = ADEP)
#Joining to ADES
fb3 <- left_join(fb2, apt_countries, by = c("ADES" = "ICAO")) %>% mutate(ADES_CTRY = CTRY, .keep = "unused", .after = ADES)

#Check  NA's
colSums(is.na(fb3))
# Very few NA's - it's safe to drop and factor now

fb4 <- fb3 %>% drop_na() %>% mutate(ADEP_CTRY = as.factor(ADEP_CTRY), ADES_CTRY = as.factor(ADES_CTRY))

#NOTE: Here we can adjust the European countries that comprises EU region by editing the "eur_countries" vector:
# Currently in european countries: "GB" "AD" "ES" "AL" "XK" "AT" "BA" "BE" "BG" "IS" "BY" "UA" "CH" "CZ" "SK" "DE" "RU" "DK" "HR" "EE" "FI" "GG" "JE" "IM" "NL" "IE" "FO" "LU" "NO" "PL" "PT" "SE" "LV" "LT" "FR" "GR" "HU" "IT" "LI" "SI" "MT" "MC" "RO" "TR" "MD" "MK" "GI" "RS" "ME" "SM" "GE" "VA"

base_dataset <- fb4 %>% mutate(ADEP_REG = as.factor(case_when(ADEP_CTRY == "US" ~ "US",
                                    ADEP_CTRY == "BR" ~ "BR",
                                    ADEP_CTRY %in% eur_countries ~ "EU",
                                    TRUE ~ "Other")), .after = ADEP_CTRY) %>%
                mutate(ADES_REG = as.factor(case_when(ADES_CTRY == "US" ~ "US",
                                    ADES_CTRY == "BR" ~ "BR",
                                    ADES_CTRY %in% eur_countries ~ "EU",
                                    TRUE ~ "Other")), .after = ADES_CTRY)
colSums(is.na(base_dataset))
# No NA's - yaaayy!!
glimpse(base_dataset)
summary(base_dataset)

```

```{r}
#quick peek
base_dataset
#
#
#
#NOW IT'S TIME TO LOOP TO OTHER MONTHS AND BIND THEM ALL!
#
#
#
```

```{r}
#quick first look
temp1 <- base_dataset %>% group_by(DATE, ADEP_REG, ADES_REG) %>% summarize(FLIGHTS = n(), .groups = "drop") %>%
  filter(ADEP_REG %in% c("BR", "EU"), ADES_REG  %in% c("BR", "EU")) %>%
  mutate(ROUTE = paste(ADEP_REG, ADES_REG, sep = "-"), .keep = "unused", .before = "FLIGHTS") %>%
  pivot_wider(names_from = ROUTE, values_from = FLIGHTS) 
temp1

#First stab at different levels of traffic

n <- 0.5
temp1 %>% ggplot(aes(x = DATE)) +
  geom_point(aes(y = `EU-EU`, color = `EU-EU` > quantile(`EU-EU`[month(DATE) %in% 1:2], probs = n)), shape = 4) +
  geom_point(aes(y = `BR-BR`, color = `BR-BR` > quantile(`BR-BR`[month(DATE) %in% 1:2], probs = n)), shape = 1) +
  labs(y  = "Flights") + 
  theme(legend.position = "bottom")

#Normalized by the median of the last 3 months of the dataset
temp2 <- temp1 %>% pivot_longer(cols = 2:5, names_to = "ROUTE", values_to = "FLIGHTS") %>% group_by(ROUTE) %>% mutate(MOVING_MEDIAN = quantile(FLIGHTS[month(DATE) %in% month(last(DATE))-2:month(last(DATE))], probs = 0.5), NORM_FLTS = FLIGHTS/MOVING_MEDIAN)
temp2

temp2 %>% ggplot(aes(x = DATE)) +
  geom_line(aes(y = NORM_FLTS, color = ROUTE)) +
  theme_minimal()
```

```{r}

```

1.2.2

@@

@@

@@

>>>>

---
title: "Assessing the Global COVID19 Impact on Air Transport with Open Data"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
# failed to render and I could not finally track what kills the pdf
#  bookdown::pdf_book:
#    base_format: rticles::ieee_article
  bookdown::word_document2
bibliography: "DASC2021.bib"
affiliation:
  ## use one only of the following
  # author-columnar: true         ## one column per author
  institution-columnar: true  ## one column per institution  
  # wide: true          ## one column wide author/affiliation fields
  institution:
    - name: EUROCONTROL
      department: Performance Review Unit
      location: Brussels (Belgium)
      #mark: 1
      author:
        - name: Rainer Koelle
          email: rainer.koelle@eurocontrol.int
    - name: DECEA
      department: Subdepartment of Operations 
      location: Rio de Janeiro (Brazil)
      #mark: 2
      author:
        - name: Fabio Lourenco Carneiro Barbosa
          email: barbosaflcb@fab.mil.br
abstract: |
  This paper approaches the impact of the pandemic as a massive service disruption of the pre-pandemic global     connectivity and regional air transport networks. In particular, the project aims to provide data analytical evidence for policy success and transformation of the air transportation system. As an aspirational goal, the industry aims to recover in a “greener” manner. The project builds on openly available data sets. The paper will be produced in a reproducible manner making the data, code, and its processing available to interested reseachers and practitioners. The open assessment will provide policy makers with a tool to assess the reaction to local or regional measures.
  <!-- motivation -->
  <!-- short background - why important -->
  <!-- approach chosen / research method -->
  <!-- experimental work -->
  <!-- results obtained -->
  <!-- conclusions / key take-away --> 
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(stats)
library(zoo)

# for corona data
# install.packages("coronavirus"), but do not load it!
# then run: coronavirus::update_dataset() 
```

```{r, echo=FALSE, message=FALSE}
## set bookdown specs/defaults for high quality output
knitr::opts_chunk$set(
   echo = FALSE, message = FALSE, warning = FALSE    # messages
  , dpi = 320               #, dev.args = list(type = "cairo") #
  , dev = "ragg_png"        # works better than cairo
#  , fig.retina = 2         # !! throws error in reticles IEEE
#  , fig.align = "center"   # !! throws error in reticles IEEE
  , fig.show = "hold"          # positioning of figure
  , fig.width = 6, fig.asp = 0.618
#  , out.width = "95%"      # !! throws error in reticles IEEE
)

@@

## load study data - as long as not outsourced to data prep
fns <- list.files(path = "./data/", pattern = "base_osn_", full.names = TRUE )
col_spec <- cols(
  ADEP = col_character(),
  ADEP_CTRY = col_character(),
  ADEP_REG = col_character(),
  ADES = col_character(),
  ADES_CTRY = col_character(),
  ADES_REG = col_character(),
  TYPE = col_character(),
  DATE = col_date(format = ""),
  CALL = col_character()
)
ds  <- fns %>% map_dfr(.f = ~ read_csv(.x))
```

```{r}
# CURRENT COMMENTS AND TO-DO'S
# I HAVE DOWNLOADED 3 FILES FOR NOW (APR/19, APR/2020, APR/2021), JUST TO START "TIDYING" AND EXPLORING.
# IT'S IN THE DATA-RAW FOLDER (NOT SHARED WITH GITHUB), AS ALWAYS.
```

@@

On March 11, 2020, the World Health Organisation has declared the novel coronavirus (COVID-19) outbreak a global pandemic.
The COVID-19 crisis posed a completely different, unexpected, and inverted challenge.
Communicable disease control resulted in massive restrictions on international and regional air traffic and passenger travel. 
Demand for air transport dropped as low as 90% of the previous "normal" in many places. Where the lack of capacity was previously the issue, now the lack of demand threatened the ATM system stability. 
In the financial perspective, airlines and airports had to deal with an unprecedented decrease in incomes. As a result, air navigation providers collected less fees for their services, due to significantly fewer flights. 
In the operational perspective, pilots and air traffic controllers practiced less. The problems and obstacles developed into many other dimensions.

@@

# # Purpose Statement
# 
# ???The purpose of this research is to 
# investigate additional dimensions in which 
# resilience could be measured, in addition
# to the current framework of delay analysis.
# 
# Research Question
# 
# ???How can we enlarge the concept of 
# resilience, so that it is applicable to
# scenarios of low traffic? 
#
# ???Research Question: 
# ???RQ1.What was the impact of the pandemic on ATM resilience?
# ???  RQ1.1 How resilience can be modeled in a low-demand scenario?
# ???  RQ1.2 How resilient were different ATM systems worldwide?
```

@@

The unprecedented decline in air traffic demand resulted in severe financial strains on the air transport industry. 
The financial support or lack thereof for airlines and airports has been widely covered in the media. 
A variety of studies showed the interplay or consequences of the travel constraints. 
To date, lower attention was given to the inherent change in terms of air transport services
Hence, the current scenario is a proper moment to further investigate the concept of resilience.

@@

> data-analytical approach - using open data / freely available (tbd: validated against organisational data)

This paper utilises a data-driven approach.
Despite the fact that aviation is a data rich environment, operational data on the previous and current traffic levels is not consistently available. 
The global traffic evaluation will be based on data from a community fed sensor network, Opensky-Network. 
<<CHECK THIS>> The operational flight data will be enriched with other publicly available dataset supporting the description of the level of the COVID pandemic and national/regional responses to the pandemic.

```{r} 
# 
#   ???RQ1.1 = through a qualitative analysis of previous proposed models
#   ???RQ1.2 = through a quantitative analysis of open data
```

@@

  + novel methodology to describe disruption, transformation, and recovery phenomena of the air transport network <<stakeholder operations>>

@@

On 11. March 2020, the World Health Organisation reacted to the steadily increasing of infections and global spread of a newly detected Corona-Virus by declaring a pandemic.
As an immediate response, many governments reacted by imposing controls to curb the further spread of communicable disease.
Next to social distancing and recommended hygiene measures, this resulted in massive restrictions on international and regional air traffic and passenger travel.

<!-- TODO: add diagram about increasing COVID infections/deaths and/or potential waves0 -->
```{r}
# coronavius (data) package is installed and updated
# call it with coronavirus:: notation
# for this paper, we reduce to May 2021 as update() will expand in future runs
covid <- coronavirus::coronavirus %>% 
  as_tibble() %>%
  filter(date <= as.POSIXct("2021-06-01"))

global_cov <- covid %>% 
  group_by(type, date) %>%
  summarise(total_cases = sum(cases), .groups = "drop") %>%
  pivot_wider(names_from = type, values_from = total_cases) %>%
  arrange(date) %>%
  mutate(active = confirmed - death - recovered) %>%
  mutate(active_total = cumsum(active),
                recovered_total = cumsum(recovered),
                death_total = cumsum(death))
```

(ref:fig-global-covid) Reported number of cases (22. Jan 2020 through 27. May 2021) [@coronavirus_data] based on data from John Hopkins University

```{r fig-global-covid, fig.cap="(ref:fig-global-covid)"}
tmp <- global_cov %>%
  select(date, contains("total")) %>%
  pivot_longer(cols = 2:4, names_to = "category", values_to = "counts")

tmp %>%

ggplot(aes(x = date, y = counts, fill = category)) +
  geom_area() +
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  theme(legend.position = c(0.2, 0.3)) +
  scale_fill_manual( name = "cases"
                    ,values = c("orange","red", "green")
                    ,labels = c("active", "death","recovered")) +
  scale_y_continuous(labels = scales::label_number_si())
```

@@

Resilience is a well-researched topic. The term is used by a diverse set of domains, communities, and research areas. In consequence there exists an abundance of theoretical definitions in the literature and numerous concepts that emerge from theses definitions and expert domains.

The term originates from physics and is nowadays used with reference to safety, security, environment and ecological systems, mental health/psychology, biological system, and others. Despite the varying contexts, across all of these disciplines the concept of resilience is closely related with the capability and ability of the focus of concern - typically the system or agent - to return to a stable state after a disruption impacted the original state.

Recently, the term is also used frequently on the political and strategic level. Policymakers, operational experts, and academes concur that the concept of resilience plays a major role when addressing and assessing the extent to which organisations and systems are prepared and capable to respond to and recover from disruptions.

Within air navigation, the term 'resilience' has been picked up by several communities, both operational and scientific.

@@

<!-- provide a pointer hook to the current interest in studying resilience (again) -->

The unprecedented decline in air transportation since March 2020 has triggered an increased interest in the topic. For example under the umbrella of the ICAO Global Air Navigation Plan Study Group, an expert team is currently working on refining the performance framework with a view to resilience. With a view to COVID, there is a dual interest in understanding resilience in air navigation/transportation:

-   political level: ADD SOME DESCRIPTION

-   operational level: ADD SOME DESCRIPTION

This paper addresses the operational dimension. Being able to characterise resilience within the operational domain will enable to address the more strategic and political decision-making.

@@

```{r osndata}
daily_tfc <- read_csv("./data/daily_osn.csv") %>% distinct()

osn_2019 <- daily_tfc %>% mutate(YR = lubridate::year(DATE)) %>%
  filter(YR == 2019) %>%
  summarise( DATE_MAX = DATE[FLTS == max(FLTS)], MAX = max(FLTS)
            ,DATE_MIN = DATE[FLTS == min(FLTS)], MIN = min(FLTS))

osn_2019_2 <- daily_tfc %>% 
  filter(DATE >= lubridate::ymd("2019-01-01") & DATE < lubridate::ymd("2020-01-01")) %>%
  summarise(MED = median(FLTS))
```

This study builds on publicly available data. Opensky Network collects crowdsourced air traffic data from more than 2500 feeders (sensor stations). To support the process of illustrating and studying the impact of the COVID pandemic on air traffic demand, a flight-by-flight dataset is provided on a monthly basis [@xavier_olive_2021_4893103]. The data spans the period since 1. January 2019.
Fig. \@ref(fig:osndaily) shows the number of daily flights tracked by Opensky Network globally.
The observed continual increase throughout the first six months in 2019 is driven by the increasing rate of ADSB equipage. Particularly, air transport operators in the United States or operating to the United States established compliance with the ADSB mandate applicable as of 1. January 2020.
In 2019, the peak daily number of tracked flights ranges just under `r pretty(osn_2019$MAX, shrink = 0.1)[2]`.
The negative spike observed on `r osn_2019$DATE_MIN` is linked to a data outage.
Throughout the second half of 2019 the median number of flights ranged just under `r pretty(osn_2019_2$MED, shrink = 0.2)[2]`.

```{r osndaily, fig.cap="(ref:osndaily)", message=FALSE}
daily_tfc %>%
  ggplot(mapping = aes(x = DATE, y = FLTS)) +
  geom_line() +
  labs(x = NULL, y = "flights")
```

@@

## Measuring Resilience

The concept of resilience (and robustness) is intimately linked with the construct of disruptions.

(ref:resiliencebasics) Resilience as a function of disturbance impact

```{r resiliencebasics, fig.cap="(ref:resiliencebasics)"}
knitr::include_graphics("figures/resilience-timeline.png")
```

 Conceptually, resilience R can be measured as the observed loss in quality of service (performance) over the time to recovery, t1-t2, for a certain level of disruption. Thus, mathematically, this represents the area covered by

$$\int_{a}^{b}[THR -LOS(t)]dt$$

as presented in Fig. \@ref(fig:resiliencebasics) (LOS: loss of service / performance, THR: associated threshold).

@@

```{r, fig.cap="test"}
#quick peek
#Read "x0_data-preparation.Rmd" first, then
head(ds)

temp <- ds %>% group_by(DATE, ADEP_REG, ADES_REG) %>% summarize(FLIGHTS = n(), .groups = "drop") %>%
  filter(ADEP_REG %in% c("BR", "EU"), ADES_REG  %in% c("BR", "EU")) %>%
  mutate(ROUTE = paste(ADEP_REG, ADES_REG, sep = "-"), .keep = "unused", .before = "FLIGHTS") %>%
  pivot_wider(names_from = ROUTE, values_from = FLIGHTS) 
#head(temp)

#First Look
temp1 <- ds %>% group_by(DATE, ADEP_REG, ADES_REG) %>% summarize(FLIGHTS = n()) %>% mutate(ROUTE = paste(ADEP_REG, ADES_REG, sep = "-"), .keep = "unused")
#head(temp1)

temp1 %>% ggplot(aes(x = DATE)) +
  geom_line(aes(y = (FLIGHTS), color = ROUTE))
```

@@

This paper addressed the identification and description of disruptions, associated transformation, and recovery phenomena of the air transport network.
The methodology will be useful to support the monitoring of the recovery levels across the globe.
It complements the toolbox for policy makers and strategic planers to determine priority measures in support of the anticipated air travel demand recovery with the roll-out and increasing levels of vaccination.

@@

<!-- final words -->
As part of this project, the requirements for an open data based flight table for the evaluation of the global air transportation network are refined. 
The results and insights of this work will be shared with the wider effort of the ICAO GANP expert group on performance.

@@

<!-- think about providing pointer to paper repo/data vs IEEE/DASC copyright reqs
The paper is written as a fully reproducible paper based on open datasets. Reproducibility and transparency will be further ensured by utilizing an open source software environment. In that respect, interested researchers or practitioners will be able to validate all aspects of the code, data, and published results. 
-->
